---
title: "Data Processing Notes"
output: html_document
date: '2022-07-13'
---
## Intro to dplyr

dplyr is an R package in the tidyverse. We can load the packae using
```{r}

# load required libraries
library(dplyr)
?dplyr

library(ggplot2)

#preview dataset
head(starwars)

```

### filtering data

we can filter data using 'filter()'

```{r}

filter(starwars, hair_color == "blond" | hair_color == "blonde")

filter(starwars, hair_color == "blond", homeworld == "Tatooine")

filtereddata = filter(starwars, hair_color == "blond", homeworld == "Tatooine", mass > 75, eye_color != "green")

filter(starwars, hair_color %in% c("blond", "brown"))

importantPlanets = c("Tatooine", "Naboo", "Hoth", "Alderaan")

filter(starwars, homeworld %in% importantPlanets)

filter(starwars, name < "Mace")

```

Tips:

* Be sure to spell the category right!
* R is case sensitive!
* Use a "==" if necessary

```{r}

head(diamonds) # first six rows
class(diamonds$cut) # specific type of data
summary(diamonds$cut) # count of each category or quartiles if numeric
filter(diamonds, price < 400)
filter(diamonds, cut > "Very Good")

unique(starwars$eye_color)

#set an order through a vector, sets up a hierarchy
factor(starwars$eye_color,
       c("red", "orange", "gold", "yellow", "green", "blue", "black"),
       ordered = TRUE)

filter(starwars, height < 100, species != "Droid")

filter(starwars, birth_year >= 100, birth_year <= 200)
filter(starwars, between(birth_year, 100, 200))

filter(starwars, mass > 100, height < 185)

```
### Missing values

```{r}
NA > 5 # it doesn't know (missing, unknown value)

mass = NA
mass == 10
is.na(mass)

x = 5
is.na(x)

NA != NA
#because...
Mary_Age = NA
Paul_Age = NA
Mary_Age == Paul_Age


filter(starwars, mass!= 75 | is.na(mass), name < "Mace")

filter(starwars, is.na(hair_color))

```

### Arranging data
'arrange()' reodrers rows, it does not remove any rows.
NA values always at the end of the order

```{r}
#lowest to highest birth year
arrange(starwars, birth_year)

# highest to lowest birth year
arrange(starwars, desc(birth_year))

arrange(starwars, name, hair_color, height)


arrange(starwars, desc(height))
arrange(starwars, height)

arrange(starwars, name)

arrange(starwars, desc(is.na(height)), desc(is.na(mass)), desc(is.na(hair_color)), desc(is.na(skin_color)), desc(is.na(eye_color)), desc(is.na(birth_year)))

arrange(starwars, desc(is.na(starwars)))

```

```{r}
#You can anonymize data

select(starwars, mass, hair_color)

select(starwars, hair_color:eye_color)

select(starwars, -(hair_color:eye_color))
select(starwars, -name)

starwars_anon = select(starwars, -name)

select(starwars, contains("_"))
select(starwars, starts_with("s"), ends_with("color"))
select(starwars, starts_with("s") & ends_with("color"))

?select

select(starwars, homeworld, species)

select(starwars, hair_color, eye_color, skin_color, birth_year)
select(starwars, ends_with("color"), birth_year)
select(starwars, hair_color:birth_year)
select(starwars, contains("_"))

```

```{r}

starwars_small = select(starwars, height, mass, birth_year)
head(starwars_small)

#new column
starwars_small = mutate(starwars_small, height_m = height/100)

#alter existing column
#non-reversible, to return to original data rerun initialization of starwars_small
mutate(starwars_small, height = height/100)

mutate(starwars_small, BMI = mass/(height_m^2))


mutate(starwars_small, 
       height_m = height/100, 
       BMI = mass/(height_m^2))


## aggregate functions, same number for all characters
mutate(starwars_small, tot_mass = sum(mass, na.rm = TRUE))
mutate(starwars_small, proportional_m = mass/sum(mass, na.rm = TRUE))

```

## summarizing and grouping data

summarize() collapses a data frame/entire column to a single row

```{r}

mutate(starwars, mean_mass = mean(mass, na.rm = T))

summarize(starwars, mean_mass = mean(mass, na.rm = T),
          mean_height = mean(height, na.rm = T))

mean(starwars$mass, na.rm = T)

species_height = summarize(group_by(starwars, species), 
          mean_height = mean(height, na.rm = T),
          count = n())

arrange(species_height, desc(count))

# the pipe operator %>%
# function(x, y) == x %>% function(y)
# don't save as an object using pipes
# when using dplyr functions, generally always start with a dataset

starwars %>%
  group_by(species) %>%
  summarize(height_mean = mean(height, na.rm = T),
            count = n()) %>%
  arrange(desc(height_mean))

```

### Sampling a designated number of rows (random)

sample_n() allows us to samplea random number of rows from out dataset

```{r}

sample_n(starwars, 10)

nrow(starwars) #total number of rows (87)
sample_frac(starwars, 0.1) #samples 10% of dataset (9)

```

Take a slice of our dataset using slice()

```{r}

starwars %>%
  group_by(species) %>%
  sample_n(1)

slice(starwars, 50:60) #slice by index value (row number)

slice_head(starwars, n = 10) #first 10 rows by index
slice_tail(starwars, n = 10) #last 10 rows by index

slice_min(starwars, order_by = mass, n = 10) #rows with lowest 10 masses

# replaces instances of "-99" with "NA"
# hero_data%Height = sub(-99, NA, hero_data$Height)

```

# subsetting data: filter/select/mutate/arrange/summarize/group_by/sample/slice
# create a visualization using new subset

grassTypes = filter(pokedata, Type.1 == "Grass" | Type.2 == "Grass", Attack > 80 | Sp..Atk > 80)
arrange(grassTypes, desc(Total))

nrow(grassTypes)

ggplot(data = grassTypes, mapping = aes(x = Attack, y = Sp..Atk, color = Generation)) +
  geom_point()


